---
title: "Activity: Bird Species"
author: "Madeline Mucciolo"
date: "June 7 2023"
output: html_notebook
---


# Set Up

```{r}
# Load `tidyverse` and `dcData` packages
library(tidyverse)
library(dcData)

# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package
data('OrdwayBirds', package = 'dcData')

# Show a few commands used to inspect the data set a few different ways
summary(OrdwayBirds)
head(OrdwayBirds)
sample_n(OrdwayBirds, 10)
nrow(OrdwayBirds)
?OrdwayBirds
```
- After briefly inspecting the data set, we see that there close to 16,000 cases and 26 variables for each. The data contains information about birds that were captured and released at the Katharine Ordway Natural History Study Area. The data contains variables describing the date and time they were captured (ex: 'Day', 'Month', 'CaptureTime'), describing the birds (ex: 'SpeciesName', 'Age', 'Sex') and other variables about the capture (ex: 'Weather', 'DataEntryPerson').


# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

Before we begin, the book gives us some instructions to select a few key variables and clean up the date formatting.  Use the commands given in the book, but be sure to study the functions used so you know what is happening.  In this case, we select `SpeciesName`, `Month`, and `Day`. Then a `mutate()` command converts each variable to character, and then converts the resulting "character string" to a number.

```{r}
# Get the data table & clean up dates (see Data Computing eBook)
OrdwayBirds <-
  OrdwayBirds %>%
  select( SpeciesName, Month, Day ) %>%
  mutate( Month = as.numeric(as.character(Month)), 
          Day = as.numeric(as.character(Day))) 

# Inspect resulting data
head(OrdwayBirds)
sample_n(OrdwayBirds, 10)
nrow(OrdwayBirds)
```
- After tidying the data into a more suitable form for our uses, the resulting table only has 3 columns ('SpeciesName', 'Month', 'Day') but still contains all of the original rows.


### Step 1
```{r}
# unique species in original data (including mis-spellings)
n_distinct(OrdwayBirds$SpeciesName)

# unique species in the clean list
data('OrdwaySpeciesNames', package = 'dcData')
n_distinct(OrdwaySpeciesNames$SpeciesNameCleaned)
```
- In the species list with incorrect spellings, there were 275 different listed species. However, in the cleaned list, there were only 109 species. 


### Step 2

```{r}
# inspect OrdwaySpeciesNames
summary(OrdwaySpeciesNames)
nrow(OrdwaySpeciesNames)
head(OrdwaySpeciesNames)

# inner join
Corrected <-
  OrdwayBirds %>%
  inner_join( OrdwaySpeciesNames ) %>% 
  select( Species = SpeciesNameCleaned, Month, Day ) %>%
  na.omit()  # cleaned up the missing ones

# inspect the data after the join
nrow(Corrected)
head(Corrected)
sample_n(Corrected, 10)
n_distinct(Corrected$Species)

```
- After joining the two tables, the new Corrected table ensures that all species names are uniformly spelled. The variable SpeciesName was used to match cases and inner join the two tables. The variable that is added is a Species variable, which contains the corresponding SpeciesNameCleaned from the OrdwaySpeciesNames table.


### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order
Corrected %>%
  group_by(Species) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(10) # a hint in the text book recommends displaying the top 10 species to choose our threshold

# define a sightings threshold; produce a table with only birds belonging to major species
Majors <- 
  Corrected %>%
  group_by(Species) %>%
  summarise(count = n()) %>%
  filter(count > 1000)

head(Majors)
```
-  I began by finding the number of bird captures for each species. I arranged these numbers into descending order, and was able to locate the top ten most captures birds during the data collection. For my definition of a major species, it is any species with count over 1000.I used this definition to filter out species into the new Majors data.  


### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
ByMonth <-
  Majors %>%
  left_join(Corrected) %>%
  select(Species, Month) %>%
  group_by(Species, Month) %>%
  summarise(Count_Per_month = n())

print(ByMonth)
```
- I created ByMonth, a table containing only the major species (as defined above) and the number of sightings for each per month. I had to use a join between the Majors table and the Corrected table.


Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)


```{r}
# ggplot commands

ByMonth %>% 
  ggplot(aes(x = Month, y = Count_Per_month, color = Species)) + 
  geom_line()

ByMonth %>% ggplot(aes(x=Month,y=Count_Per_month ,fill=Species)) +
  geom_bar(stat='identity',position='stack', width=.9) 

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?  
- American Gold Finch and Black-capped Chickadee are the only two major species present year-round.  


##### 2. Which species are migratory, that is, primarily present in one or two seasons?
- Field Sparrows, Slate-colored Juncos and Tree Swallows are migratory species.


##### 3. What is the peak month for each major species? (bullet list)
- The peak month for American Goldfinch is the 10th month (October). The peak month for Black-capped Chickadee is the 11th month (November). The peak month for Field Sparrow is the 5th month (May). The peak month for Slate-colored Juno is the 10th month (October). The peak month for Tree Swallow is the 3rd month (March).



##### 4. Which major species are seen in good numbers for at least 6 months of the year?
- All of the major species are seen in good numbers for at least 6 months of the year.

